---
title: "Mutiara_35236043_Assignment3"
output: html_document
name: "Mutiara Hernowo"
studentID: "35236043"
date: "2025-10-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(123)
```

## Library
Load required packages for data manipulation, modeling, and visualization
```{r library}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(tibble)
library(randomForest)
library(xgboost)
```

## Data Preparation
Import and prepare raw datasets from BPS (productivity) and BMKG (climate)
### Read file
Import CSV files containing rice productivity and climate data
Sources: BPS (Badan Pusat Statistik/Indonesian Bureau of Statistics) and BMKG (Indonesian Bureau of Meteorology, Climatology, and Geophysics)
```{r data_analysis}
# Read file paths
climate <- read_csv("all_climate_data.csv", na = c("N/A"))
rice <- read_csv("rice_field_2019_2024.csv")
```

### Rename column
Standardize column names for consistency and readability
Convert Indonesian names to English for international reproducibility
```{r rename_column}
# Rename column name in each data to make processing easier
## Climate data
colnames(climate) <- tolower(colnames(climate))

## Field data
rice <- rice %>%
  rename(
    `year` = Year,
    `province` = Province,
    `yield_area` = `Yield Area`,
    `field_prod` = `Field Productivity`,
    `rice_prod` = Production
  )

summary(climate)
summary(rice)
```

## Data Preprocessing
### Missing Value
Identify and handle missing values using appropriate imputation strategies
Strategy: Provincial mean imputation to preserve spatial patterns
```{r missing_value}
# Check Missing Value
climate_missing <- climate %>%
  group_by(province, year) %>%
  summarise(
    missing_temp_max = sum(is.na(tx)),
    missing_temp_avg = sum(is.na(tavg)),
    missing_rainfall = sum(is.na(rr)),
    missing_humidity = sum(is.na(rh_avg)),
    missing_sunshine = sum(is.na(ss)),
    .groups = "drop"
  )
climate_missing

# Long form
missing_long <- climate_missing %>%
  select(
    province,
    year,
    starts_with("missing_")
  ) %>%
  gather(
    key = "indicator",
    value = "missing_value",
    -province, -year
  ) %>%
  mutate(
    indicator = recode(
      indicator,
      missing_temp_avg = "temp_avg",
      missing_temp_max = "temp_max",
      missing_rainfall = "rainfall",
      missing_humidity = "humidity",
      missing_sunshine = "sunshine"
    )
  )

# Plot Missing Values
## Plot Rainfall because Rainfall has many missing values
missing_rainfall <- missing_long %>%
  filter(indicator == "rainfall")

ggplot(missing_rainfall, aes(x = factor(year), y = province, fill = missing_value)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = missing_value), 
            color = "black", 
            size = 4.5,
            fontface = "bold") +
  scale_fill_gradient(low = "#fff5f0", high = "#b30000",
                      name = "Missing\nCount") +
  labs(title = "Missing Values: Rainfall", 
       x = "Year", 
       y = NULL) +
  theme_linedraw() +
  theme(
    axis.text.y = element_text(size = 13, hjust = 1),  # FIX OVERLAP!
    axis.text.x = element_text(size = 13),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 18, hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(face = "bold")
  )

## The other indicators
missing_others <- missing_long %>%
  filter(indicator != "rainfall")

ggplot(missing_others, aes(x = factor(year), y = province, fill = missing_value)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = missing_value), 
            color = "black", 
            size = 2,
            fontface = "bold") +
  facet_wrap(~ indicator, ncol = 2) +
  scale_fill_gradient(low = "#fff5f0", high = "#b30000",
                      name = "Missing\nCount") +
  labs(title = "Missing Values: Climate Indicators", 
       x = "Year", 
       y = NULL) +
  theme_linedraw() +
  theme(
    axis.text.y = element_text(size = 8, hjust = 1),
    axis.text.x = element_text(size = 8),
    strip.text = element_text(face = "bold", size = 10, color = "white"),
    strip.background = element_rect(fill = "#404040", color = "black"),
    panel.grid = element_blank(),
    panel.spacing = unit(1.5, "lines"),
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8, face = "bold")
  )
```

### Data Cleaning
The heatmaps show the missing value in used variable and I use imputation to handle this missing value.
```{r missing_value_handling}
# Data cleaning
# Change `DATE` data type
climate$date <- dmy(climate$date)

# Handle missing value using imputation
climate <- climate %>%
  group_by(province, year) %>%
  mutate(
    rr = ifelse(is.na(rr), median(rr, na.rm = TRUE), rr),
    tavg = ifelse(is.na(tavg), mean(tavg, na.rm = TRUE), tavg),
    tx = ifelse(is.na(tx), mean(tx, na.rm = TRUE), tx),
    rh_avg = ifelse(is.na(rh_avg), mean(rh_avg, na.rm = TRUE), rh_avg),
    ss = ifelse(is.na(ss), mean(ss), ss)
  ) %>%
  ungroup()

# Check missing value after cleaning
climate_clean <- climate %>%
  summarise(
    na_temp = sum(is.na(tavg)),
    na_temp_max = sum(is.na(tx)),
    na_rainfall = sum(is.na(rr)),
    na_humidity = sum(is.na(rh_avg)),
    na_sunlight = sum(is.na(ss))
  )

colSums(is.na(climate_clean))
climate
```

### Aggregation: Annual data
Aggregate monthly/daily data to annual level for alignment with harvest cycles
Note: If data already annual, this step validates no within-year duplicates
```{r annual_aggregation}
annual_climate <- climate %>%
  group_by(province, year) %>%
  summarise(
    `rain_sum` = sum(rr, na.rm = TRUE),
    `temp_max_avg` = mean(tx, na.rm = TRUE),
    `temp_avg` = mean(tavg, na.rm = TRUE),
    `hot_days` = sum(tx > 34, na.rm = TRUE),
    `humid_avg` = mean(rh_avg, na.rm = TRUE),
    `sunshine_sum` = sum(ss, na.rm = TRUE),
    .groups = "drop"
  )
annual_climate
summary(annual_climate)
```

## Exploratory Data Analysis
### Descriptive Analysis
Summary statistics for all variables: mean, median, sd, min, max, quartiles
```{r descriptive_analysis}
climate_rice <- annual_climate %>%
  left_join(rice, by = c("province", "year"))

# Descriptive Analysis
province_summary <- climate_rice %>%
  group_by(province) %>%
  summarise(
    # Rice metrics
    avg_field_productivity = mean(field_prod, na.rm = TRUE),
    sd_field_productivity = sd(field_prod, na.rm = TRUE),
    avg_yield_area = mean(yield_area, na.rm = TRUE),
    avg_rice_production = mean(rice_prod, na.rm = TRUE),
    
    # Climate metrics
    avg_hot_days = mean(hot_days, na.rm = TRUE),
    avg_temp_max = mean(temp_max_avg, na.rm = TRUE),
    avg_rainfall = mean(rain_sum, na.rm = TRUE),
    avg_humidity = mean(humid_avg, na.rm = TRUE),
    
    # Sample size
    n_years = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_field_productivity))

print(province_summary)
```

### Distribution Analysis
Visualize variable distributions to assess normality, skewness, and outliers.
Purpose: Inform normalization decisions and validate data quality
```{r dist_analysis_climate}
# Rainfall
ggplot(climate_rice, aes(x = rain_sum)) +
  geom_histogram(bins = 20, fill = "#3498DB", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(rain_sum, na.rm = TRUE)), 
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Annual Rainfall",
       x = "Rainfall (mm)", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

# Temperature (Max)
ggplot(climate_rice, aes(x = temp_max_avg)) +
  geom_histogram(bins = 20, fill = "#E74C3C", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(temp_max_avg, na.rm = TRUE)),
             color = "blue", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Maximum Temperature",
       x = "Max Temperature (°C)", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

# Hot Days
ggplot(climate_rice, aes(x = hot_days)) +
  geom_histogram(bins = 20, fill = "#F39C12", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(hot_days, na.rm = TRUE)),
             color = "blue", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Hot Days",
       x = "Hot Days (n)", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

# Humidity
ggplot(climate_rice, aes(x = humid_avg)) +
  geom_histogram(bins = 20, fill = "#1ABC9C", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(humid_avg, na.rm = TRUE)),
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Humidity",
       x = "Humidity (%)", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

# Sunshine
ggplot(climate_rice, aes(x = sunshine_sum)) +
  geom_histogram(bins = 20, fill = "#F1C40F", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(sunshine_sum, na.rm = TRUE)),
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Sunshine Duration",
       x = "Sunshine (hours)", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

```{r dist_analysis_rice}
# Yield Area
ggplot(climate_rice, aes(x = yield_area)) +
  geom_histogram(bins = 20, fill = "#9B59B6", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(yield_area, na.rm = TRUE)),
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Yield Area",
       x = "Yield Area (ha)", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

# Field Productivity
ggplot(climate_rice, aes(x = field_prod)) +
  geom_histogram(bins = 20, fill = "#2ECC71", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(field_prod, na.rm = TRUE)),
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Field Productivity",
       x = "Productivity (ku/ha)", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

# Rice Production
ggplot(climate_rice, aes(x = rice_prod)) +
  geom_histogram(bins = 20, fill = "#E67E22", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(rice_prod, na.rm = TRUE)),
             color = "red", linetype = "dashed", size = 1) +
  labs(title = "Distribution of Rice Production",
       x = "Production (tons)", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

```
### Time Series Trends
Quantify relationships between climate variables and rice productivity
```{r temp_max_trend}
ggplot(annual_climate, aes(x = year, y = temp_max_avg, color = province, group = province)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2019:2024) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Maximum Temperature Trend (2019-2024)",
    subtitle = "Average across all provinces",
    x = "Year",
    y = "Max Temperature (°C)",
    color = "Province"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    legend.title = element_text(face = "bold")
  )
```
```{r temp_avg_trend}
ggplot(annual_climate, aes(x = year, y = temp_avg, color = province, group = province)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2019:2024) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Average Temperature Trend (2019-2024)",
    subtitle = "Average across all provinces",
    x = "Year",
    y = "Average Temperature (°C)",
    color = "Province"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    legend.title = element_text(face = "bold")
  )
```

```{r humidity_trend}
ggplot(annual_climate, aes(x = year, y = humid_avg, color = province, group = province)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2019:2024) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Average Humidity Trend (2019-2024)",
    subtitle = "Average across all provinces",
    x = "Year",
    y = "Average Humidity (%)",
    color = "Province"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    legend.title = element_text(face = "bold")
  )
```
```{r rainfall_trend}
ggplot(annual_climate, aes(x = year, y = rain_sum, color = province, group = province)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2019:2024) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Rainfall Trend (2019-2024)",
    subtitle = "Total across all provinces",
    x = "Year",
    y = "Total Rainfall (mm)",
    color = "Province"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    legend.title = element_text(face = "bold")
  )
```
```{r sunshine_trend}
ggplot(annual_climate, aes(x = year, y = sunshine_sum, color = province, group = province)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2019:2024) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Sunshine Trend (2019-2024)",
    subtitle = "Total across all provinces",
    x = "Year",
    y = "Total Sunshine (hours)",
    color = "Province"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    legend.title = element_text(face = "bold")
  )
```
```{r hot_days_trend}
ggplot(annual_climate, aes(x = year, y = hot_days, color = province, group = province)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 2019:2024) +
  scale_color_brewer(palette = "Paired") +
  labs(
    title = "Total Hot Days Trend (2019-2024)",
    subtitle = "Total across all provinces",
    x = "Year",
    y = "Total Hot Days (days)",
    color = "Province"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 11),
    legend.title = element_text(face = "bold")
  )
```
Based on the time trend analysis, there are some patterns in 2020-2021 where the COVID-19 occurs. The temperature, rainfall, humidty, and sunshine duration tend to going up, while the hot days going down. It can be a signal of good quality air during pandemic.

### Correlation Analysis
Method: Pearson (linear) and Spearman (non-linear) correlation coefficients
```{r correlation_analysis}
corr_data <- climate_rice %>%
  select(
    # Climate predictors
    rain_sum, temp_max_avg, temp_avg, hot_days, humid_avg, sunshine_sum,
    # Rice outcomes
    yield_area, field_prod, rice_prod
  ) %>%
  drop_na()

# Pearson Correlation
pearson_corr <- cor(corr_data, method = "pearson")
pearson_pval <- cor.mtest(corr_data, method = "pearson")$p
print(round(pearson_corr, 3))
pearson_pval

# Spearman Correlation
spearman_corr <- cor(corr_data, method = "spearman")
spearman_pval <- cor.mtest(corr_data, method = "spearman")$p
print(round(spearman_corr, 3))
spearman_pval

# CREATE HEATMAP
climate_vars <- c("rain_sum", "temp_max_avg", "temp_avg", "hot_days", 
                  "humid_avg", "sunshine_sum")
rice_vars <- c("yield_area", "field_prod", "rice_prod")

# Function to prepare data
prep_heatmap <- function(cor_matrix, pval_matrix, method_name) {
  cor_matrix[climate_vars, rice_vars] %>%
    as.data.frame() %>%
    rownames_to_column("climate") %>%
    gather(
      key = "rice",
      value = "correlation",
      -climate
    ) %>%
    left_join(
      pval_matrix[climate_vars, rice_vars] %>%
        as.data.frame() %>%
        rownames_to_column("climate") %>%
        gather(
          key = "rice",
          value = "pval",
          -climate
        ),
      by = c("climate", "rice")
    ) %>%
    mutate(
      method = method_name,
      sig = case_when(
        pval < 0.001 ~ "***",
        pval < 0.01 ~ "**",
        pval < 0.05 ~ "*",
        TRUE ~ ""
      ),
      climate = recode(climate,
                       rain_sum = "Rainfall",
                       temp_max_avg = "Max Temp",
                       temp_avg = "Mean Temp",
                       hot_days = "Hot Days",
                       humid_avg = "Humidity",
                       sunshine_sum = "Sunshine"),
      rice = recode(rice,
                    yield_area = "Yield Area",
                    field_prod = "Field Productivity",
                    rice_prod = "Rice Production")
    )
}

# Prepare both
pearson_data <- prep_heatmap(pearson_corr, pearson_pval, "Pearson")
spearman_data <- prep_heatmap(spearman_corr, spearman_pval, "Spearman")

# Combine
both_data <- bind_rows(pearson_data, spearman_data)

# Plotting
ggplot(both_data, aes(x = rice, y = climate, fill = correlation)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = sprintf("%.2f", correlation)), 
            size = 3, vjust = -0.3) +
  geom_text(aes(label = sig), size = 3, vjust = 1.5) +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                       midpoint = 0, limits = c(-1, 1)) +
  facet_wrap(~method, ncol = 2) +
  labs(title = "Climate-Rice Correlations: Pearson vs Spearman",
       x = NULL, y = NULL,
       caption = "*** p<0.001, ** p<0.01, * p<0.05",
       fill = "Correlation") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    strip.text = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid = element_blank(),
    legend.title = element_text(face = "bold")
  )
```
## Data Integration and Feature Engineering
### Phenology data
Creating a phenology data because the phenology-specific data (planting dates, growth stages) not available.
Adaptation: Use annual climate aggregates as phenology-proxy indicators
Justification: Annual metrics capture seasonal stress patterns on rice growth
F0: rainfall, temperature, hot days frequency, humidity, sunshine
F1: seasonal data. Indonesia has two season: wet and dry season and each of them has rain and hot days. Next, this will be assigned using F0 feature.
F2: flowering windows month when the rice ususally enter flowering phase
```{r features}
features <- climate %>%
  mutate(month = month(date)) %>%
  group_by(province, year) %>%
  summarise(
    # ---- F0: annual aggregates ----
    rain_sum     = sum(rr),
    temp_avg    = mean(tavg, na.rm = TRUE),
    hot_days     = sum(tx > 34, na.rm = TRUE),
    humid_avg   = mean(rh_avg, na.rm = TRUE),
    sunshine_sum = sum(ss, na.rm = TRUE),

    # ---- F1: seasonal (wet/dry) ----
    wet_rain = sum(rr[month %in% c(10,11,12,01,02,03)], na.rm = TRUE),
    dry_rain = sum(rr[month %in% 04:09], na.rm = TRUE),
    wet_hot  = sum((month %in% c(10,11,12,01,02,03)) & tx > 34, na.rm = TRUE),
    dry_hot  = sum((month %in% 04:09) & tx > 34, na.rm = TRUE),

    # ---- F2: flowering windows ----
    hot_flowering_wet = sum((month %in% 01:03) & tx > 34, na.rm = TRUE),  # January - March: Flowering Days during Wet Season
    hot_flowering_dry = sum((month %in% 06:08) & tx > 34, na.rm = TRUE),  # June - August Flowering Days during Dry Season
    .groups = "drop"
  ) %>%
  mutate(
    hot_days_flowering = pmax(hot_flowering_wet, hot_flowering_dry)  # use max temperature of both season
  ) %>%
  select(-hot_flowering_wet, -hot_flowering_dry)
```

### Join dataset
Merge climate and productivity data by province-year keys
Result: unified data for modelling
```{r target_join}
data <- field %>%
  left_join(features, by = c("province", "year"))

# write_csv(data, "climate_metadata.csv")
data
```

### Create feature set F0, F1, F2
Progressive feature sets to evaluate incremental value of climate variables
- F0: base climate
- F1: F0 + seasonal
- F2: F1 + phenology
```{r}
make_features <- function(feature_data, set = c("F0", "F1", "F2")) {
  set <- match.arg(set) # defining the set of each features
  base <- c("rain_sum","temp_avg","hot_days","sunshine_sum","humid_avg")
  seas <- c("wet_rain","dry_rain","wet_hot","dry_hot")
  flow <- c("hot_days_flowering")
  feats <- switch(set,
                  F0 = base,
                  F1 = c(base, seas),
                  F2 = c(base, seas, flow))
  feature_data %>%
    select(province, year, field_prod, all_of(feats)) %>%
    drop_na()
}

# Apply function
dF0 <- make_features(data, "F0")
dF1 <- make_features(data, "F1")
dF2 <- make_features(data, "F2")

# Extract feature names
fF0 <- setdiff(names(dF0), c("province","year","field_prod"))
fF1 <- setdiff(names(dF1), c("province","year","field_prod"))
fF2 <- setdiff(names(dF2), c("province","year","field_prod"))

# Verify
cat("F0 features:", paste(fF0, collapse = ", "), "\n")
cat("F1 features:", paste(fF1, collapse = ", "), "\n")
cat("F2 features:", paste(fF2, collapse = ", "), "\n")
```
## Method and Evaluation
### LOYO Evaluation
Leave-One-Year-Out cross-validation: train on n-1 years, test on 1 year
Purpose: Assess temporal generalization (mimics operational deployment)
```{r loyo_evaluation}
# Root Mean Absolute Percentage
rmape <- function(y, yhat){
  mean(abs((y - yhat) / y)) * 100
}

# Evaluation using Leave-One-Year-Out (LOYO)
eval_loyo <- function(feature_data, feats, model = c("ols","rf","xgb")) {
  model <- match.arg(model)
  years <- sort(unique(data$year))               # year list for LOYO

  map_df(years, function(y) {
    tr <- feature_data %>% filter(year != y)            # train: all year without y
    te <- feature_data %>% filter(year == y)            # test : only y
    fml <- as.formula(paste("field_prod ~", paste(feats, collapse = " + ")))
    
    if (model == "ols") {
      fit  <- lm(fml, data = tr)               # baseline linear
      pred <- predict(fit, te)
    } else if (model == "rf") {
      fit  <- randomForest(fml, data = tr,     # non-linear for tabular
                           ntree = 500,
                           mtry  = floor(sqrt(length(feats))))
      pred <- predict(fit, te)
    } else {
      Xtr <- as.matrix(tr[, feats])
      ytr <- tr$field_prod
      Xte <- as.matrix(te[, feats])
      yte <- te$field_prod

      dtr <- xgb.DMatrix(data = Xtr, label = ytr)
      dte <- xgb.DMatrix(data = Xte)

      fit <- xgboost(
        data = dtr,
        objective = "reg:squarederror",
        nrounds = 300,
        max_depth = 5,
        eta = 0.1,
        subsample = 0.8,
        colsample_bytree = 0.8,
        min_child_weight = 1,
        gamma = 0,
        verbose = 0
      )
      pred <- predict(fit, dte)
    }

    tibble(
      year_left_out = y,
      RMSE = sqrt(mean((te$field_prod - pred)^2)),   # original scale (ton/ha)
      MAPE = rmape(te$field_prod, pred)              # error (%)
    )
  }) %>%
    summarise(                                   
      RMSE_mean = mean(RMSE, na.rm = TRUE),
      MAPE_mean = mean(MAPE, na.rm = TRUE),
    )
}
```

### Model Testing
Evaluate OLS, Random Forest, and XGBoost across F0, F1, F2 feature sets
Objective: Empirically identify optimal model-feature combination
```{r}
# Testing using different models
# OLS
eval_f0_ols <- eval_loyo(dF0, fF0, "ols")
eval_f1_ols <- eval_loyo(dF1, fF1, "ols")
eval_f2_ols <- eval_loyo(dF2, fF2, "ols")

# RF
eval_f0_rf <- eval_loyo(dF0, fF0, "rf")
eval_f1_rf <- eval_loyo(dF1, fF1, "rf")
eval_f2_rf <- eval_loyo(dF2, fF2, "rf")

# XGBoost
eval_f0_xgb <- eval_loyo(dF0, fF0, "xgb")
eval_f1_xgb <- eval_loyo(dF1, fF1, "xgb")
eval_f2_xgb <- eval_loyo(dF2, fF2, "xgb")
```

### Result Comparison
Compare all models and identify best performer
Criteria: Lowest RMSE (primary), lowest MAPE (secondary)
``` {r result_comparison}
result <- tibble(
  Model   = rep(c("OLS","RF","XGBoost"), each = 3),
  `Feature Set` = rep(c("F0","F1","F2"), times = 3),
  RMSE    = c(eval_f0_ols$RMSE_mean,
              eval_f1_ols$RMSE_mean,
              eval_f2_ols$RMSE_mean,
              eval_f0_rf$RMSE_mean,
              eval_f1_rf$RMSE_mean,
              eval_f2_rf$RMSE_mean,
              eval_f0_xgb$RMSE_mean,
              eval_f1_xgb$RMSE_mean,
              eval_f2_xgb$RMSE_mean),
  MAPE    = c(eval_f0_ols$MAPE_mean,
              eval_f1_ols$MAPE_mean,
              eval_f2_ols$MAPE_mean,
              eval_f0_rf$MAPE_mean,
              eval_f1_rf$MAPE_mean,
              eval_f2_rf$MAPE_mean,
              eval_f0_xgb$MAPE_mean,
              eval_f1_xgb$MAPE_mean,
              eval_f2_xgb$MAPE_mean)
)

result_round <- result %>%
  mutate(
    across(c(RMSE,MAPE), ~round(.x, 2))
  )
result_round
write_csv(result_round, "result.csv")

# Plot the result
plot_df <- result_round %>%
  mutate(
    Model = factor(Model, levels = c("OLS","RF","XGBoost")),
    `Feature Set` = factor(`Feature Set`, levels = c("F0","F1","F2"))
  ) %>%
  pivot_longer(cols = c(RMSE, MAPE), names_to = "Metric", values_to = "Value")

best_df <- plot_df %>%
  group_by(Metric) %>%
  filter(Value == min(Value)) %>%
  ungroup()

ggplot(plot_df, aes(x = `Feature Set`, y = Value, fill = Model)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  geom_text(aes(label = round(Value, 2)),
            position = position_dodge(width = 0.75),
            vjust = -0.3, size = 2.5) +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(title = "Model Performance by Feature Set",
       x = "Feature Set", y = "Score (lower is better)") +
  theme_linedraw(base_size = 12) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
    )
```

### Save Model
Train final model on full dataset and save for operational deployment
Model: Best performer from comparison (typically XGBoost-F2)
```{r save_model}
# Save best model
if (!dir.exists("models")) dir.create("models")

# Train final model on all data
X_all <- as.matrix(dF2[, fF2])
y_all <- dF2$field_prod
d_all <- xgb.DMatrix(data = X_all, label = y_all)

# Final XGBoost
final_xgb <- xgboost(
        data = d_all,
        objective = "reg:squarederror",
        nrounds = 300,
        max_depth = 5,
        eta = 0.1,
        subsample = 0.8,
        colsample_bytree = 0.8,
        min_child_weight = 1,
        gamma = 0,
        verbose = 0
        )

xgb.save(final_xgb, "models/xgboost_f2_best.model")
saveRDS(fF2, "models/features.rds")
```

## Early Warning System
Operational deployment: Apply trained model to current year (2024) for risk alerts
```{r}
# Load model
best_model <- xgb.load("models/xgboost_f2_best.model")
features <- readRDS("models/features.rds")

# Calculate Baseline (Historical Average)
baseline <- dF2 %>%
  filter(year < 2024) %>%
  group_by(province) %>%
  summarise(baseline_prod = mean(field_prod, na.rm = TRUE))

# Using 2024 for simulation
year_2024 <- dF2 %>%
  filter(year == 2024) %>%
  select(province, all_of(features))

# Prediction matrix
prediction_matrix <- as.matrix(year_2024[, features])
prediction <- predict(best_model, prediction_matrix)

# Compare and Classify Risk
alert <- year_2024 %>%
  mutate(
    predicted_prod = prediction,
    province = province
  ) %>%
  left_join(baseline, by = "province") %>%
  mutate(
    change_in_productivity = (predicted_prod - baseline_prod) / baseline_prod * 100,
    risk = case_when(
      change_in_productivity < -10 ~ "HIGH RISK",
      change_in_productivity < -5 ~ "MODERATE RISK",
      TRUE ~ "NORMAL")
  ) %>%
  arrange(change_in_productivity)

# Show Early Warning Alert
print(alert %>% select(province, predicted_prod, baseline_prod, change_in_productivity, risk))

# Plot Alert
ggplot(alert, aes(x = reorder(province, change_in_productivity), y = change_in_productivity, fill = risk)) +
  geom_col() +
  geom_hline(yintercept = -10, linetype = "dashed", color = "red") +
  coord_flip() +
  scale_fill_manual(values = c("HIGH RISK" = "#E74C3C", 
                                "MODERATE RISK" = "#F39C12",
                                "NORMAL" = "#2ECC71")) +
  labs(title = "Early Warning: Predicted Yield Productivity Change (2024)",
       x = NULL, y = "Change from Baseline (%)", fill = "Risk Level") +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )
```


